---
title: "Spatial"
author: "Jeff Matheson"
output: 
  html_notebook:
    toc: true
    toc_float: true
---

# Introduction

These are my notes and code snippets relating to creation of maps using r.

## Main packages for mapping in R
```{r}
library(sp)         ## Key package. Though being replaced by sf.
library(sf)         ## Replaces sp. sp still needed for some things.
library(rgdal)      ## Bindings for GDAL(?) seems to provide helper functions

library(raster)     ## This is key for raster manipulation.
library(rasterVis)
library(maptools)   ## Haven't used this yet

library(maps)       # Collection of maps

library(rgeos)      # ? Has some data. Geometry Engine
library(tmap)
library(tmaptools)
library(ggmap)      ## Spatial Visualization with ggplot2
library(maps)

library(bcmaps)
```

```{r}
## For dynamic maps

library(leaflet) # Key package for dynamic maps. 
library(mapview) # Wrapper for leaflet. Great for quick viewing.

## Web integration

library(htmlwidgets) 
library(shiny) 
library(webshot)
```
## Static Maps

`gglot` keeps thingsin the tidyverse.

`rnaturalearth` is good source for data and works nice with sf and ggplot.

```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")
world %>% 
  ggplot() +
  geom_sf() +
  ggtitle("World")

```


# Leaflet
```{r}
m <- leaflet() %>% 
      addTiles() 

setView(lng=-120.912, lat=56.1938, zoom=16)
m

OpenTopoMap  
Esri.WorldStreetMap  
Wikimedia ## good for roads  
m %>% addProviderTiles(providers$Esri.WorldImagery)  
m %>% addProviderTiles(providers$Esri.NatGeoWorldMap)  
m %>% addProviderTiles(providers$Stamen.Watercolor)  

t <- leaflet(footprint)
t

m <- leaflet(options = leafletOptions(minZoom = 5, maxZoom = 30)) %>% 
      addTiles(group = "OSM (default)")  %>%   
      addProviderTiles(providers$Esri.WorldStreetMap, group = "Street Map") %>%    
      addProviderTiles(providers$Wikimedia, group = "Wikimedia") %>%
      addProviderTiles(providers$Esri.WorldImagery, group = "Imagery") %>%
      addRectangles(
        lng1=-122.032, lat1=55.85,
        lng2=-120.64, lat2=56.37,
        fillColor = "transparent"
        ) %>%
# addMarkers of addPopups  
      addMarkers(lng=-120.912, lat=56.1938, popup="Site C", labelOptions(sticky=FALSE)) %>%
      # Layers control
      addLayersControl(
        baseGroups = c("OSM (default)", "Street Map", "Wikimedia", "Imagery"),
        options = layersControlOptions(collapsed = FALSE))

m

# Use this to set an initial view. 
m %>% setView(m, lng=-120.912, lat=56.1938, zoom=8)

# Add measurement tool
m %>% addMeasure(primaryLengthUnit = "metres",
                 primaryAreaUnit = "hectares")

# Add minimap
m %>% addMiniMap(tiles = providers$Wikimedia,
                toggleDisplay = TRUE)
```




Lat/long NAD83 is EPSG 4269 Lat/long WGS84 is EPSG 4326 UTM NAD83 zone 10N is EPSG 26911

How to reproject

    newcrs = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
    footprint <- spTransform(footprint, newcrs)

Read/Write shapefile library(rgdal) writeOGR(footprint, dsn = ".", "fp\_wgs84", driver="ESRI Shapefile")

Read UTM cooridnates and write same datafile with lat/long dat \<- read\_csv("WOODto2019.csv") dat \<- SpatialPointsDataFrame(coords = cbind(woodat$UTM\_Northing, woodat$UTM\_Easting), data = woodat, proj4string=CRS("+proj=utm +zone=10 +north +datum=NAD83"))  
dat \<- spTransform(wooutm, CRS("+proj=longlat +datum=WGS84 +no\_defs +ellps=WGS84 +towgs84=0,0,0")) datout \<- (cbind([dat\@data](mailto:dat@data), [dat\@coords](mailto:dat@coords))) %\>% rename(Longitude = coords.x1, Latitude = coords.x2)

Rasters GDALinfo("") raster file attributes before reading.


# Google maps API 
key = AIzaSyDq1pZGYfF1PmyxxvWvLPW22TH7dAOkt90

```{r}
register_google(key = "AIzaSyDq1pZGYfF1PmyxxvWvLPW22TH7dAOkt90", write = TRUE)
has_google_key()
google_key()
hdf <- get_map("houston, texas")
ggmap(hdf, extent = "normal")
ggmap(hdf) # extent = "panel", note qmap defaults to extent = "device"
ggmap(hdf, extent = "device")

```
# Shiny

```{r}
library(shiny)

ui <- fluidPage()
server <- function(input, output){}
shinyApp(ui=ui, server = server)


<<<<<<< HEAD
```
=======
Rasters
GDALinfo("") raster file attributes before reading.  


# Read File Geodatabase with sf
And then save as KML.  
```{r}

library(sf)
library(mapview)

cliff_swallows <- read_sf("C:/Users/jeff.matheson/Downloads/deerfoot_cliff/d4781c57ea5c4ad79a6a7717e3c8f9f5.gdb")
c("C:/Users/jeff.matheson/Downloads/deerfoot_cliff/d4781c57ea5c4ad79a6a7717e3c8f9f5.gdb")

mapview(cliff_swallows)

st_write(cliff_swallows, 
         driver = "KML", 
         "C:/Users/jeff.matheson/Downloads/deerfoot_cliff/CLSW.kml")


```

>>>>>>> ff6e6677cbd265d9aafdc1644381420e905a5990
