---
title: "Diversity and Community Analyses"
author: "Jeff Matheson"
output: 
  html_notebook:
    toc: true
    toc_float: true
  github_document: default
  
---

# Biodiversity Analyses

library(vegan)
library(dismo)                Species distribution models
library(BiodiversityR)
library(indicspecies)         Indicator species analysis


RankAbun.1 <- rankabuncomp(stxsp, y=st.year, factor='Survey.Name', scale='abundance', 
                           legend=F, ylim = c(0,600), rainbow=T)
RankAbun.1 <- rankabuncomp(stxsp, y=st.year, factor='Survey.Name', scale='proportion',
                           legend=T, rainbow=T)
RankAbun.1 <- rankabuncomp(stXspAbunSMSA, y=stXyearSMSA, factor='Survey.Name', scale='proportion',
                           legend=T, rainbow=T)
```{r}
library(iNEXT)
out <- iNEXT(spider, q=c(0,1,2), datatype="abundance", size=m)
ggiNEXT(x, type=1, se=TRUE, facet.var="order", color.var="site", grey=FALSE)

DataInfo(spXyearSMSA)
out <- iNEXT(spXyearSMSAn, q=0, datatype="incidence_freq",endpoint=336)
out <- iNEXT(spXyearSMSAn, q=c(0,1,2), datatype="incidence_freq",endpoint=336)
ggiNEXT(out,type=1,se=TRUE)
ggiNEXT(out,type=2,se=TRUE)
ggiNEXT(out, type=1, facet.var="order")

library(SpadeR)
SimilarityMult(spxyear.1i, datatype = "incidence_freq", q = 0, nboot = 200)
```
---

# Dissimilarity
Two key functions:
vegdist(), method - "hellinger"
decostand()


# Markdown

kable() for presentation of tables.

---
# Ordination
Putting things in order.
* Arrange samples samples along gradients
* Map data to lower dimensions

## Unconstrained
First we look for major variation, then relate to environmental variation. 

PCA: Linear model. Most useful for env data, sometimes species data.
CA: Unimodal method. Species data and presence-absence, esp. where non-linear responses are observed. Very similar to PCA, ecept weighted form. 
PCO and NMDS: Any kind of data. Doesn't care about gradinet, just dissimilarity. 

## NMDS from using Palmer Penguins data. 

```{r}
pengv <- penguins %>% 
  select(3:6)

nmds <- metaMDS(pengv, distance = "bray")

# extract NMDS scores (x and y coordinates)
data.scores = as.data.frame(scores(nmds))

# add columns to data frame 
data.scores$species <- penguins$species
data.scores$sex <- penguins$sex

n_plot <- ggplot(data.scores, aes(x = NMDS1, y = NMDS2)) + 
  geom_point(size = 2, aes(shape = sex, colour = species))+ 
  theme(axis.text.y = element_text(colour = "black", size = 10, face = "bold"), 
        axis.text.x = element_text(colour = "black", face = "bold", size = 10), 
        legend.text = element_text(size = 10, face ="bold", colour ="black"), 
        legend.position = "right", axis.title.y = element_text(face = "bold", size = 11), 
        axis.title.x = element_text(face = "bold", size = 10, colour = "black"), 
        legend.title = element_text(size = 11, colour = "black", face = "bold"), 
        panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
        legend.key=element_blank()) + 
  labs(x = "NMDS1", colour = "Species", y = "NMDS2", shape = "Sex") 
n_plot

ggsave(plot = n_plot, "nmds_plot.png")
```
