---
title: "Spatial"
author: "Jeff Matheson"
output: 
  html_notebook:
    toc: true
    toc_float: true
  github_document: default
---

# Mapping

## Main packages for mapping in R
library(sp)         #Key package. Though being replaced by sf.
library(sf)         #Replaces sp. sp still needed for some things.
library(rgdal)      #Bindings for GDAL(?) seems to provide helper functions
library(raster)     #This is key for raster manipulation.
library(rasterVis)
library(maptools)   #Haven't used this yet

library(rgeos)      ? Has some data. Geometry Engine
library(tmap)
library(tmaptools)
library(ggmap)      #Spatial Visualization with ggplot2

library(rspatial)

library(maps)

library(bcmaps)

## For dynamic maps

library(leaflet)    Key package for dynamic maps. 
library(mapview)    Wrapper for leaflet. Great for quick viewing.

## Web integration 
library(htmlwidgets)
library(shiny)
library(webshot)

## Static Maps
{gglot} keeps thingsin the tidyverse.

{rnaturalearth} is good source for data and works nice with sf and ggplot.

```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")
world %>% 
  ggplot() +
  geom_sf() +
  ggtitle("World")
```

Mapview, good for quick view
`mapview(x)`

Leaflet
```
m <- leaflet() %>% 
  addTiles() 
```

Lat/long NAD83 is EPSG 4269
Lat/long WGS84 is EPSG 4326
UTM NAD83 zone 10N is EPSG 26911 

How to reproject
```
newcrs = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
footprint <- spTransform(footprint, newcrs)
```

Read/Write shapefile
library(rgdal)
writeOGR(footprint, dsn = ".", "fp_wgs84", driver="ESRI Shapefile")

Read UTM cooridnates and write same datafile with lat/long 
dat <- read_csv("WOODto2019.csv") 
dat <- SpatialPointsDataFrame(coords = cbind(woodat$UTM_Northing, woodat$UTM_Easting), 
                              data = woodat, proj4string=CRS("+proj=utm +zone=10 +north +datum=NAD83"))  
dat <- spTransform(wooutm, CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
datout <- (cbind(dat@data, dat@coords)) %>% 
                  rename(Longitude = coords.x1, Latitude = coords.x2)

Rasters
GDALinfo("") raster file attributes before reading.  


# Read File Geodatabase with sf
And then save as KML.  
```{r}

library(sf)
library(mapview)

cliff_swallows <- read_sf("C:/Users/jeff.matheson/Downloads/deerfoot_cliff/d4781c57ea5c4ad79a6a7717e3c8f9f5.gdb")
c("C:/Users/jeff.matheson/Downloads/deerfoot_cliff/d4781c57ea5c4ad79a6a7717e3c8f9f5.gdb")

mapview(cliff_swallows)

st_write(cliff_swallows, 
         driver = "KML", 
         "C:/Users/jeff.matheson/Downloads/deerfoot_cliff/CLSW.kml")


```

